---
title: "Beyond the Report Card: What are the insights at the county level that we can observe?"
author: "Thu Dang, Nathaniel Reimer, Jeremy Hubinger"
execute: 
  echo: false
  warning: false
  message: false
bibliography: Library.bib
---

```{r packages}
library(tidyverse)
library(MetBrewer)
library(sf)
library(rnaturalearth)
library(classInt) #install.packages('classInt')
library(ggspatial) #install.packages('ggspatial')
library(ggthemes) #install.packages('ggthemes')
library(RColorBrewer)
library(patchwork)
library(leaflet)
library(probably) #install.packages('probably')
library(glmnet)
library(tidymodels)
tidymodels_prefer()
```

```{r color}
palname = "Homer2"
defaultcolor = met.brewer(palname,10)[9]
```

```{r load datasets}
load("CombinedDataLongFormat.RData")
load("WideMergedData.RData")
```

```{r load cleaned dataset for county achievement metrics}
math_ca_gcs_df <- read_csv("AllDatasets/math_ca_gcs_df.csv")
rla_ca_gcs_df <- read_csv("AllDatasets/rla_ca_gcs_df.csv")
```

```{r maps load shapefile}
ca_counties <- read_sf('CA_Counties')
ca_counties <- ca_counties %>%
  st_transform(crs = "NAD83")
```

# An overall look at Academic Achievement across Californian counties

In this analysis, we define academic achievement as the difference between the Grade Cohort Standardize (GCS) Scores and NAEP standard. As a reminder of the interpretation for GCS scores, we can look at this example again: If 4th-grade students at the school of interest have a GCS value of 5.03, these students' scores indicate a level equivalent to 5th-grade, which is about one grade level higher than the national average (the reference group) in math.

To create this map, we calculated the difference in GCS scores from the grade levels for each grade-school cohort, then aggregated them by the Californian counties. We divided the score differences into 3 categories: *1-2 years behind*, *Less than 1 year behind*, and *Less than 1 year ahead*. 


```{r gcs map data cleaning}
school_map <- merged_table %>%
  select(CDSCode, School, gradecenter, gcs_mn_avg_ol, Longitude, Latitude) %>%
  filter(!is.na(gradecenter) & !is.na(gcs_mn_avg_ol)) %>% #7390 after NA gradecenter, 6997 after both conditions
  distinct %>%
  mutate(diff_from_gradecenter = gcs_mn_avg_ol - gradecenter) %>%
  mutate(gcs_type = ifelse(diff_from_gradecenter < 0, "Below NAEP Average", ifelse(diff_from_gradecenter > 0, "Above NAEP Average", "At NAEP Average")))

school_map_sf <- school_map %>% 
  st_as_sf(coords = c('Longitude', 'Latitude'), crs = "NAD83")

school_county_sf <- st_join(school_map_sf, ca_counties) 

grade_diff_by_county <- school_county_sf %>%
  st_drop_geometry() %>% #removes geometry - makes calculation more efficient 
  group_by(NAMELSAD) %>%
  summarise(mean_gcs_diff = mean(diff_from_gradecenter))

grade_diff_by_county_geo <- ca_counties %>%
  left_join(grade_diff_by_county, by="NAMELSAD")
```

```{r gcs map}
grade_diff_by_county_with_breaks_geo <- grade_diff_by_county_geo %>%
  mutate(grade_diff_cat = factor(cut(mean_gcs_diff, breaks=c(-2, -1, 0, 1, 2))))

pal <- colorFactor(
  palette = met.brewer(palname,3)[1:3],
  domain = grade_diff_by_county_with_breaks_geo$grade_diff_cat
)

leaflet(grade_diff_by_county_with_breaks_geo) %>%
  addPolygons(
    fillColor = ~pal(grade_diff_cat),
    opacity = .25,
    color = "white",
    fillOpacity = .5,
    group = "County",
    popup = ~paste0(NAMELSAD, 
                    "<br>Grade Distance from National Average: ", round(mean_gcs_diff, digits = 3))
  )  %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
    addLayersControl(
    baseGroups = "County",
    position = "topright"

  ) %>%
  addLegend(colors = met.brewer(palname,3)[1:3], values = ~grade_diff_cat, title = NULL, labels = c('1-2 years behind', 'Less than 1 year behind', 'Less than 1 year ahead'), className = "info legend County")

```

<br>
Looking at this map, it is clear that there is a significant difference in academic performance across various regions in California. The Bay Area and Orange County appear to outperform other counties, along with affluent tourist counties like Placer and El Dorado. However, for the remaining counties, while approximately half of them exhibit students with academic performance on par with the national average, a significant proportion of counties have students lagging behind by 1-2 years. 

We recognize that there may be a huge wealth disparity among diverse communities in California. Therefore, our next step is to delve deeper into the social factors that may be contributing to what we're observing on this map. 

  
# Further exploration into the county-level math and reasoning language arts (RLA) achievement of different student groups

Indeed, according to Cano and Hong, California faces challenging income disparities and intricate demographics that extend beyond its public education system. Since 2008, California has implemented extensive reforms aimed at allocating more resources to high-needs students and addressing educational disparities, with some success. However, despite these efforts, Black, Latino, and low-income students continue to significantly underperform compared to Asian American, white, and wealthier students in both math and reading proficiency [@cal_matter]. A UCLA research shows that Californian schools are the most segregated for Latinos, with 58% attending schools that are highly segregated. Additionally, The study also revealed that over 50% of Black students in California are concentrated in only 25 out of 1,000 school districts [@ucla_article].

Since we're interested in how different socioeconomic backgrounds might influence education achievement, we looked into how students classified as economically disadvantaged according to the Californian standard perform compared to those in the non-economically disadvantaged group in terms of GCS scores. By definition, students are typically considered economically disadvantaged if they come from a low-income household or meet other criteria for poverty, such as eligibility for free or reduced-price meals, homelessness, or foster care. The criteria for California can be found [here](https://documentation.calpads.org/Glossary/AccountabilitySubgroupData/Socio-EconomicallyDisadvantagedSubgroup/#socio-economically-disadvantaged-subgroup). The GCS score differences are broken down into 6 categories, ranging from being 2+ years behind to 2+ years ahead compared to the NAEP standard. The GCS categories are also divided into two disciplines, i.e. math and RLA for both economically disadvantaged and non-economically disadvantaged groups.


```{r math data cleaning for ecd and nec leaflet}
summary_math_data <- ca_counties %>%
  left_join(math_ca_gcs_df, by=c("NAMELSAD" = "sedacountyname")) %>%
  filter(student_category %in% c('ecd','nec')) %>%
  filter(!is.na(gcs_value) & !is.na(student_category)) %>%
  select(NAMELSAD, grade, year, gcs_value, student_category, clean_student_category) %>%
  mutate(gcs_diff = gcs_value - grade) %>%
  st_drop_geometry() %>%
  group_by(NAMELSAD, student_category, clean_student_category) %>%
  summarise(mean_gcs_diff = mean(gcs_diff)) %>%
  ungroup()

summary_math_data_with_geo <- ca_counties %>%
  select(NAMELSAD) %>%
  left_join(summary_math_data, by="NAMELSAD")

summary_math_data_with_geo <- summary_math_data_with_geo %>%
  mutate(mean_gcs_diff_cat = factor(cut(mean_gcs_diff, breaks=c(-3, -2, -1, 0, 1, 2, 3))))
```

```{r rla data cleaning for ecd and nec leaflet}
summary_rla_data <- ca_counties %>%
  left_join(rla_ca_gcs_df, by=c("NAMELSAD" = "sedacountyname")) %>%
  filter(student_category %in% c('ecd','nec')) %>%
  filter(!is.na(gcs_value) & !is.na(student_category)) %>%
  select(NAMELSAD, grade, year, gcs_value, student_category, clean_student_category) %>%
  mutate(gcs_diff = gcs_value - grade) %>%
  st_drop_geometry() %>%
  group_by(NAMELSAD, student_category, clean_student_category) %>%
  summarise(mean_gcs_diff = mean(gcs_diff)) %>%
  ungroup()

summary_rla_data_with_geo <- ca_counties %>%
  select(NAMELSAD) %>%
  left_join(summary_rla_data, by="NAMELSAD")

summary_rla_data_with_geo <- summary_rla_data_with_geo %>%
  mutate(mean_gcs_diff_cat = factor(cut(mean_gcs_diff, breaks=c(-3, -2, -1, 0, 1, 2, 3))))
```

```{r math ecd performance map}
pal <- colorFactor(
  palette = met.brewer(palname,6)[1:6],
  domain = summary_math_data$mean_gcs_diff_cat
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
    addLayersControl(
    baseGroups = "County",
    position = "topright"
  ) %>%
  addPolygons(
    data = summary_math_data_with_geo %>% filter(student_category == 'ecd'),
    fillColor = ~pal(mean_gcs_diff_cat),
    opacity = .25,
    color = "white",
    fillOpacity = .5,
    group = "Math: Economically Disadvantaged",
    popup = ~paste0(NAMELSAD, 
                    "<br>Grade Distance from National Average: ", round(mean_gcs_diff, digits = 3)
    )
  )  %>% 
  addPolygons(
    data = summary_math_data_with_geo %>% filter(student_category == 'nec'),
    fillColor = ~pal(mean_gcs_diff_cat),
    opacity = .25,
    color = "white",
    fillOpacity = .5,
    group = "Math: Non-Economically Disadvantaged",
    popup = ~paste0(NAMELSAD, 
                    "<br>Grade Distance from National Average: ", round(mean_gcs_diff, digits = 3)
    )
  )  %>% 
  addPolygons(
    data = summary_rla_data_with_geo %>% filter(student_category == 'ecd'),
    fillColor = ~pal(mean_gcs_diff_cat),
    opacity = .25,
    color = "white",
    fillOpacity = .5,
    group = "RLA: Economically Disadvantaged",
    popup = ~paste0(NAMELSAD, 
                    "<br>Grade Distance from National Average: ", round(mean_gcs_diff, digits = 3)
    )
  )  %>% 
  addPolygons(
    data = summary_rla_data_with_geo %>% filter(student_category == 'nec'),
    fillColor = ~pal(mean_gcs_diff_cat),
    opacity = .25,
    color = "white",
    fillOpacity = .5,
    group = "RLA: Non-Economically Disadvantaged",
    popup = ~paste0(NAMELSAD, 
                    "<br>Grade Distance from National Average: ", round(mean_gcs_diff, digits = 3)
    )
  )  %>% 
  addLayersControl(baseGroups = c("Math: Economically Disadvantaged", "Math: Non-Economically Disadvantaged", "RLA: Economically Disadvantaged", "RLA: Non-Economically Disadvantaged"),
                   position = "bottomleft", options = layersControlOptions(collapsed = FALSE)) %>% 
  addLegend(colors = met.brewer(palname,6)[1:6], values = ~mean_gcs_diff_cat, title = NULL, labels = c('2+ years behind','1-2 years behind', 'Less than 1 year behind', 'Less than 1 year ahead', '1-2 years ahead', '2+ years ahead'), group = "Economically Disadvantaged")
```

<br>
Although it is intuitive that economically disadvantaged students perform worse than their non-economically disadvantaged peers, it is still not at all less shocking when we looked at this map for the first time. The general trend shows that at county level, there is almost no overlapping in academic achievement between economically disadvantaged and non-economically disadvantaged student groups, for both math and RLA. Specifically, while the economically disadvantaged group is lagging behind by 1 to 2+ years behind the national standard, the more privileged group performs much better, having a performance gap of less than 1 year to being 2+ years ahead of the national standard.



