---
title: "Beyond the Report Card: Let's discuss some results!"
author: "Thu Dang, Nathaniel Reimer, Jeremy Hubinger"
execute: 
  echo: false
  warning: false
  message: false
---

```{r packages}
library(tidyverse)
library(MetBrewer)
library(sf)
library(rnaturalearth)
library(classInt) #install.packages('classInt')
library(ggspatial) #install.packages('ggspatial')
library(ggthemes) #install.packages('ggthemes')
library(RColorBrewer)
library(patchwork)
library(leaflet)
library(probably) #install.packages('probably')
library(glmnet)
library(tidymodels)
tidymodels_prefer()
```

```{r color}
palname = "Homer2"
defaultcolor = met.brewer(palname,10)[9]
```

```{r load datasets}
load("CombinedDataLongFormat.RData")
load("WideMergedData.RData")
```

```{r load cleaned dataset for county achievement metrics}
math_ca_gcs_df <- read_csv("AllDatasets/math_ca_gcs_df.csv")
rla_ca_gcs_df <- read_csv("AllDatasets/rla_ca_gcs_df.csv")
```

```{r maps load shapefile}
ca_counties <- read_sf('CA_Counties')
ca_counties <- ca_counties %>%
  st_transform(crs = "NAD83")
```

# What did we find, and what do they mean?

- A paragraph summarizing county- and school-level insights

- A paragraph on adjusting funding and how important it is to be mindful of that or else they'll be misinformation

- A paragraph on LASSO results and talk about the big picture of the results

- Growth score is pretty robust since there are no variables that actually predict it.


# Moving forward

- What are some of the next steps? 

- What should we have done differently? 




