---
title: "Beyond the Report Card: Let's discuss some results!"
author: "Thu Dang, Nathaniel Reimer, Jeremy Hubinger"
execute: 
  echo: false
  warning: false
  message: false
---

```{r packages}
library(tidyverse)
library(MetBrewer)
library(sf)
library(rnaturalearth)
library(classInt) #install.packages('classInt')
library(ggspatial) #install.packages('ggspatial')
library(ggthemes) #install.packages('ggthemes')
library(RColorBrewer)
library(patchwork)
library(leaflet)
library(probably) #install.packages('probably')
library(glmnet)
library(tidymodels)
tidymodels_prefer()
```

```{r color}
palname = "Homer2"
defaultcolor = met.brewer(palname,10)[9]
```

```{r load datasets}
load("CombinedDataLongFormat.RData")
load("WideMergedData.RData")
```

```{r load cleaned dataset for county achievement metrics}
math_ca_gcs_df <- read_csv("AllDatasets/math_ca_gcs_df.csv")
rla_ca_gcs_df <- read_csv("AllDatasets/rla_ca_gcs_df.csv")
```

```{r maps load shapefile}
ca_counties <- read_sf('CA_Counties')
ca_counties <- ca_counties %>%
  st_transform(crs = "NAD83")
```

# What did we find, and what do they mean?

There is a high degree of variation in both the county and school-level achievement data. Wealthier counties and counties with large urban areas see higher average student achievement. When we separate economically disadvantaged and non-economically disadvantaged students the economically disadvantaged students have lower average achievement in both math and english. At the school level there is more significant variability. High-achieving and low-achieving schools can be found close to each other but there are many groups of high and low-achieving schools. Diving deeper into school funding, we find that schools with a higher proportion of disadvantaged students often have higher funding than schools with fewer disadvantaged students which may be misleading at first glance. When we account for the greater need of these economically disadvantaged students, there still appears to be a funding gap. Funding schools equally is not an equitable or effective solution to bridge the gap in socioeconomic status, leaving gaps in academic achievement unbridged.

-   A paragraph on LASSO results and talk about the big picture of the results <--------- **JEREMY YOU DO THIS ONE** 
        -   Growth score is pretty robust since there are no variables that actually predict it. <------ **AND THIS ONE LMAO**

# Limitations and moving forward

The main limitation in our analysis is that we drop observations with missing data. We also only had the time to look at LASSO for modeling. 

Moving forward, in further analysis, we should use spatial models to impute 

-   What are some of the next steps?

-   What should we have done differently? <------------------ **Thu finish this whole section**
